---
title: "Lab 7 - Hannah Pryor"
format: gfm
editor: visual
---

<br>

Load packages and data

```{r, message=FALSE}

# Load required packages
library(tidyverse)
library(knitr)
library(nycflights13)

flights |> head() |> kable()

```

<br>

Question 1: Compute the average delay by destination, then join on the `airports` data frame so you can show the spatial distribution of delays. Here’s an easy way to draw a map of the United States:

```{r, message=FALSE, warning=FALSE}
library(maps) #install.packages("maps")

airports |>
  semi_join(flights, c("faa" = "dest")) |>
  ggplot(aes(lon, lat)) +
    borders("state") +
    geom_point() +
    coord_quickmap()
```

<br>

Use the `size` or `colour` of the points to display the average delay for each airport

```{r, warning=FALSE, message=FALSE}
# add a column with average delay to the flights dataframe and pull out onlu the required columns, also rename dest to faa to make merging easier
flight_delays <- flights |> 
  group_by(dest) |> 
  mutate(avg_delay=mean(arr_delay, na.rm=T)) |> 
  select(dest, arr_delay, avg_delay) |> 
  rename(faa=dest)

# create a combined table by joining the selected columns from flights with the airports dataset
combined_table <-
  left_join(airports, flight_delays)

# create a plot colored by average delay using this combined table in ggplot
combined_table |> 
  semi_join(flights, c("faa" = "dest")) |>
   ggplot(aes(lon, lat, color=avg_delay, na.rm=T)) +
    borders("state") +
    geom_point() +
    coord_quickmap()
```

<br>

Question 2: Is there a relationship between the age of a plane and its delays?

```{r, warning=FALSE}
view(planes)

planes_flights <-
  left_join(flights, planes, by=join_by(tailnum)) |> 
  rename(flight_year=year.x, make_year=year.y) |> 
  mutate(age=flight_year-make_year)

planes_flights |> 
  ggplot(aes(age, arr_delay)) +
  geom_col()

# the newest and oldest planes are very close to on-schedule, but the planes that are inbetween have a lot of variation in the arrival time/delays. This is an interesting pattern...
```

<br>

Question 3: What weather conditions make it more likely to see a delay?

```{r}
View(weather)

flights_weather <-
  left_join(flights, weather)

flights_weather |> 
  ggplot(aes(
```
